---
title: "[Figure 6] Perform GSEA on output from DESeq code"
output:
  pdf_document: default
  html_notebook: default
---

This script takes in the output of deseq.Rmd and tximport_setup.Rmd to create a list of genes ranked by sign of shrunken fold change and p-value and perform GSEA to detect gene lists enriched at the extremes of this list.

Once GSEA is performed, you can generate a figure using EnrichmentMap. Here is an EnrichmentMap tutorial once you have your ranked lists: https://enrichmentmap.readthedocs.io/en/latest/QuickTour.html

In general, this pipeline module is based on:
https://baderlab.github.io/Cytoscape_workflows/EnrichmentMapPipeline/supplemental_protocol1_rnaseq.html

Notes on how I did GSEA:

1. Use this code to generate and save two files:
Gene ranked by a combination of p-value and sign of fold change
Pathway definition file

2. Load into GSEA

3. Click preranked analysis

Help on doing preranked analysis here:
http://software.broadinstitute.org/gsea/doc/GSEAUserGuideFrame.html?_GSEAPreranked_Page

How to see/analyze results: https://software.broadinstitute.org/gsea/doc/GSEAUserGuideFrame.html
(that also explains what you see when you click on the index)

4. Click leading edge analysis, pick folder from application cache, click Load GSEA results, sort.
FDR cutoff (q value) of 0.25 suggested appropriate by GSEA: https://software.broadinstitute.org/gsea/doc/GSEAUserGuideFrame.html


# Prepare workspace

## set heap
```{r}
# set max heap size of rJava - default is 512 which is easy to overwhelm
options(java.parameters = "-Xmx12000m")
```

## Import the things you need
```{r}

# import .csv file with user paths and info written out
userVars <- read.delim("../userVars.csv", stringsAsFactors=FALSE, sep=",")

# print(userVars) # uncomment to print for documentation which paths you are using

# user-determined variables used for generating filenames
currDate <- userVars[userVars$variable=='currentDate', 'path']
projDir <- userVars[userVars$variable=='projectDirectory', 'path']

# path to output of tximport_setup.Rmd
pathToSetupData <- userVars[userVars$variable=='date_mostRecent_setup', 'path']

# path to DESeq output
pathToDESeqOutput <- paste0(projDir,'/output/figure6/deseq/',userVars[userVars$variable=='date_mostRecent_setup', 'path'],'/other' )



```


# Load required packages and variables
```{r}
# Load edgeR
knitr::opts_knit$set(cache = TRUE)

tryCatch(expr = { library("edgeR")}, 
         error = function(e) { 
           source("https://bioconductor.org/biocLite.R")
           biocLite("edgeR")}, 
         finally = library("edgeR"))

working_dir <- paste(getwd(),"data",sep="/")

#The field in the class definition file that defines the classes of the data.
data_classes <- "SUBTYPE"

# Load DESeqOutput - RData
load(paste0(projDir, "output/figure6/deseq/", pathToSetupData, "/RData/", pathToSetupData, "_deseq_deseqOutput_INTvCLEAN_gene.RData") )

# Load other variables
load(paste0(projDir, "output/figure6/tximport_setup/", pathToSetupData, "/RData/", pathToSetupData , "_tximport_setup_setupVars.RData") )

```


## Make folders for output
```{r}

# name for files etc
expName <- 'GSEA'

# make folder to contain experiment output

## for output
pathToAllProjectDir <- paste0(projDir, 'output/figure6/', expName)
dir.create(pathToAllProjectDir)

## main folder
pathToProjectDir <- paste0(projDir, 'output/figure6/', expName,'/', currDate)
dir.create(pathToProjectDir)

## make folder to contain non-RData output
pathToProjectOutput <- paste0(pathToProjectDir,'/other')
dir.create(pathToProjectOutput)

## make folder for RData vars
pathToRData <- paste0(pathToProjectDir,'/RData')
dir.create(pathToRData)

```



## Create GSEA input list
```{r}

#create ranked list of DESeq output
ranks_RNAseq <- sign(myResults_df$log2FoldChange) * -log10(myResults_df$padj)

#gene names from the TCGA set contain gene name and entrez gene ids separated by ‘|’
# for all subsequent enrichment analysis we need to have just one id.  Separate the names 
# into their two ids.
genenames <- unlist(row.names(myResults_df))

# make ranks file
ranks_RNAseq <- cbind(genenames, ranks_RNAseq)
colnames(ranks_RNAseq) <- c("GeneName","rank")

#sort ranks in decreasing order
ranks_RNAseq <- ranks_RNAseq[order(as.numeric(ranks_RNAseq[,2]),decreasing = TRUE),]

# remove NAs
ranks_RNAseq <- ranks_RNAseq[ which(!(is.na(ranks_RNAseq[,2]))) ,]

# write table of ranks
setwd(pathToProjectOutput)
write.table(ranks_RNAseq, paste(pathToProjectOutput, paste0("DE_ranks_",expName,".rnk"),sep="/"), 
            col.name = TRUE, sep="\t", row.names = FALSE, quote = FALSE)

```


## download latest and greatest pathway definition file
```{r}
#install required R and bioconductor packages
tryCatch(expr = { library("RCurl")}, 
         error = function(e) {  install.packages("RCurl")}, 
         finally = library("RCurl"))

# set wd
setwd(pathToProjectOutput)

# url with current file
gmt_url = "http://download.baderlab.org/EM_Genesets/current_release/Human/symbol/"

#list all the files on the server
filenames = getURL(gmt_url)
tc = textConnection(filenames)
contents = readLines(tc)
close(tc)

#get the gmt that has all the pathways and does not include terms inferred from electronic annotations(IEA)
#start with gmt file that has pathways only
rx = gregexpr("(?<=<a href=\")(.*.GOBP_AllPathways_no_GO_iea.*.)(.gmt)(?=\">)",
  contents, perl = TRUE)
gmt_file = unlist(regmatches(contents, rx))

dest_gmt_file <- paste(pathToProjectOutput, paste("Supplementary_Table3_", gmt_file, sep="") ,sep="/")

download.file(
    paste(gmt_url,gmt_file,sep=""),
    destfile=dest_gmt_file
)

```

# Create paths needed for GSEA run
```{r}
# establish new paths
pathToRanks <- paste0(pathToProjectOutput, '/DE_ranks_GSEA.rnk')
pathToExpr <- paste0(pathToProjectOutput, '/EM_expressionFile_RNASeq.txt')
pathToGMT <- paste(pathToProjectOutput, paste("Supplementary_Table3_", gmt_file, sep="") ,sep="/")

# create params variable 
params <- data.frame(gsea_jar =  userVars[userVars$variable=='pathToGSEAjar', 'path'],
                     working_dir = pathToProjectOutput,
                     analysis_name = 'run_repeat_GSEA',
                     run_gsea = pathToProjectOutput,
                     rnk_file = pathToRanks,
                     expression_file = pathToExpr,
                     pval_thresh = 0.05,
                     fdr_thresh = 0.25)

```


# Load required libraries
```{r}
#install required R and bioconductor packages
tryCatch(expr = { library("RCurl")}, 
         error = function(e) {  install.packages("RCurl")}, 
         finally = library("RCurl"))

#use library
tryCatch(expr = { library("limma")}, 
         error = function(e) { source("https://bioconductor.org/biocLite.R")
           biocLite("limma")}, 
         finally = library("limma"))
tryCatch(expr = { library("Biobase")}, 
         error = function(e) { source("https://bioconductor.org/biocLite.R")
           biocLite("Biobase")}, 
         finally = library("Biobase"))
tryCatch(expr = { library("ggplot2")}, 
         error = function(e) { install.packages("ggplot2")}, 
         finally = library("ggplot2"))

#For creating json and communicating with cytoscape
tryCatch(expr = { library("httr")}, 
         error = function(e) { install.packages("httr")}, 
         finally = library("httr"))
tryCatch(expr = { library("RJSONIO")}, 
         error = function(e) { install.packages("RJSONIO")}, 
         finally = library("RJSONIO"))

```

# Configure the configurable parameters
```{r}
#path to GSEA jar 
# In order to run GSEA automatically you need to speciry the path to the gsea jar file.
gsea_jar <- params$gsea_jar

#Gsea takes a long time to run.  If you have already run GSEA manually or previously there is no need to re-run GSEA.  Make sure the 
# gsea results are in the current directory and the notebook will be able to find them and use them.
run_gsea = params$run_gsea

#navigate to the directory where you put the downloaded protocol files.
working_dir <- params$working_dir

# leave blank if you want the notebook to discover the gsea directory for itself
#gsea_directory = paste(working_dir,"Mesen_vs_Immuno.GseaPreranked.1497635459262",sep="/") 
gsea_directory = params$gsea_directory

analysis_name <- params$analysis_name
rnk_file <- params$rnk_file
expression_file <- params$expression_file


```


# RUN GSEA
Note: HAVE TO RUN JAVA 8. In Terminal, type java -version and make sure that says Java 8. If not, uninstall and re-install Java.
```{r}

for (i in 1:3) {
  print(paste0('Run ', i,  'starting...'))
  command <- paste("java  -Xmx12G -cp", gsea_jar,  
                   "xtools.gsea.GseaPreranked -gmx", 
                   dest_gmt_file, "-rnk" ,file.path(rnk_file), 
                   "-collapse false -nperm 5000 -permute gene_set -scoring_scheme weighted -rpt_label ",
                   analysis_name,
                   paste0("  -num 100 -plot_top_x 20 -rnd_seed ", substr(as.numeric(Sys.time()),6,10) , " -set_max 200 -set_min 15 -zip_report false -out" ,
                   working_dir, 
                   paste0("-gui false > gsea_output_", i, ".txt") ) ,sep=" ")
  system(command)
}


```

# Get name of output directory
```{r}
if(gsea_directory == ""){
  gsea_directories <- list.files(path = working_dir, pattern = "\\.GseaPreranked")

  #get the details on the files
  details = file.info(file.path(getwd(),working_dir,gsea_directories))
  #order according to newest to oldest
  details = details[with(details, order(as.POSIXct(mtime),decreasing = TRUE)), ]

  #use the newest file:
  gsea_output_dir <- row.names(details)[1]

} else {
  gsea_output_dir <- gsea_directory
}

```


# HOUSEKEEPING
```{r}

sessionInfo()
setwd(pathToProjectOutput)
save.image(file = paste0(currDate,'_',expName,'_allVars.RData'))

```



