---
title: "[Figure 6] Heatmap of Differentially Expressed Genes"
output: html_notebook
---

This code takes the output of tximport_setup.Rmd and deseq.Rmd. It visualizes the output as a heatmap.

# Set up workspace

To ensure that this code is usable across a variety of platforms, the user-provided variables (e.g. the date of the run, the path to data files) are stored in the .csv file "userVars", which should be in the directory containing the directory that holds the .Rmd scripts.

## Import the variables you need
```{r}
# import .csv file with user paths and info written out
userVars <- read.delim("../userVars.csv", stringsAsFactors=FALSE, sep=",")

# print(userVars) # uncomment to include output in notebook to document variables in .html notebook output for this run

# user-determined variables used for generating filenames
currDate <- userVars[userVars$variable=='currentDate', 'path']
projDir <- userVars[userVars$variable=='projectDirectory', 'path']
myAlpha <- userVars[userVars$variable=='myAlpha', 'path']

pathToSetupData <- userVars[userVars$variable=='date_mostRecent_setup', 'path']

# path to DESeq output
pathToDESeqOutput <- paste0(projDir,'/output/figure6/deseq/',userVars[userVars$variable=='date_mostRecent_setup', 'path'],'/other' )

```

## load needed variables
```{r}
# Load setup data
load( paste0(projDir, 'output/figure6/tximport_setup/', pathToSetupData, "/RData/", pathToSetupData, "_tximport_setup_setupVars.RData") )

# Load deseq data
load( paste0(pathToDESeqOutput, "/DEGenes_top100.RData") )
load( paste0(pathToDESeqOutput, "/namesOfDEGenes.RData") )


## order samples by PCR integration
sampleTable_plus <- sampleTable_plus[order(sampleTable_plus$PCR_integration),]

```


## Make folders for this
```{r}
# name for files etc
expName <- 'heatmaps'

# make folder to contain experiment output

## for output
pathToAllProjectDir <- paste0(projDir, 'output/figure6/', expName)
dir.create(pathToAllProjectDir)

## main folder
pathToProjectDir <- paste0(projDir, 'output/figure6/', expName,'/', currDate)
dir.create(pathToProjectDir)

## make folder to contain non-RData output
pathToProjectOutput <- paste0(pathToProjectDir,'/other')
dir.create(pathToProjectOutput)

## make folder for RData vars
pathToRData <- paste0(pathToProjectDir,'/RData')
dir.create(pathToRData)



```

## create dataset for plotting
```{r}
# make data frame
myData <- assay(dds_kal_vst_bc)

# make order match sampleTable
myData <- myData[, match(sampleTable_plus$DESeqAnalysisID, colnames(myData))]

# take transpose because that's required for WGCNA
plotData <- t(myData)

```


# Heatmap: DE Genes
Feed in batch corrected data: https://support.bioconductor.org/p/99480/

## prepare plotting data
```{r}

plot_de <- plotData[,which(colnames(plotData) %in% namesOfDEGenes)]
plot_de_100 <- plotData[,which(colnames(plotData) %in% row.names(deseqOutput_signif_df_100))]

```

## Plot: batch corrected vst-normalized expression data
```{r}

# load needed libraries
library(RColorBrewer)
library(DESeqAid)

# define annotation colors
annColors <- list(PCR_integration = c(N='#bdbdbd',Y='#636363'))

# plot
setwd(pathToProjectOutput)
# the way I did it in older versions
DE_heatmap(plot_de, 
           sampleTable_plus, 
           c('PCR_integration'), 
           'DESeqAnalysisID', 
           30,15, 
           'DE_expression_YlOrRed.pdf', 
           myColors = colorRampPalette(rev(brewer.pal(n = 7, name ="YlOrRd")))(100), 
           myAnnotationColors = annColors, 
           26)

DE_heatmap(plot_de_100, 
           sampleTable_plus, 
           c('PCR_integration'), 
           'DESeqAnalysisID', 
           30,15, 
           'DE_expression_top100_YlOrRed.pdf', 
           myColors = colorRampPalette(rev(brewer.pal(n = 7, name ="YlOrRd")))(100), 
           myAnnotationColors = annColors, 
           26)


```

## Plot: batch corrected vst-normalized Z-SCORES
```{r}

# swap out names for LIne IDs
row.names(plot_de) <- sampleTable_plus[which( row.names(plot_de) %in% sampleTable_plus[,'DESeqAnalysisID'] ), 'Line.x']
row.names(plot_de_100) <- sampleTable_plus[which( row.names(plot_de_100) %in% sampleTable_plus[,'DESeqAnalysisID'] ), 'Line.x']

# !! temporarily make copy of sampleTable with futz'ed names for Julien to change
sampleTable_plus_futz <- sampleTable_plus
sampleTable_plus_futz$Line.x <- as.character(sampleTable_plus_futz$Line.x)
sampleTable_plus_futz[which(sampleTable_plus_futz$DESeqAnalysisID == 'WT_L3_P26'), 'Line.x'] <- 'WT_L3_P26_NIH511.3'
sampleTable_plus_futz[which(sampleTable_plus_futz$DESeqAnalysisID == 'WT_L5_P20'), 'Line.x'] <- 'WT_L5_P20_8343.4'
sampleTable_plus_futz[which(sampleTable_plus_futz$DESeqAnalysisID == 'WT_L5_P21'), 'Line.x'] <- 'WT_L5_P21_8343.5'
sampleTable_plus_futz$Line.x <- as.factor(sampleTable_plus_futz$Line.x)


# create annotations
  my_pheno_col <- data.frame(sampleTable_plus_futz[,c('PCR_integration')])
  row.names(my_pheno_col) <- sampleTable_plus_futz[,'Line.x']
  colnames(my_pheno_col) <- c('PCR_integration')
  
# plot pheatmap
  setwd(pathToProjectOutput)
  pheatmap(plot_de, 
           annotation_row = my_pheno_col, 
           cluster_rows = FALSE, gaps_row = 26, 
           filename = 'degene_scaled_BuPu_LineNames.pdf', 
           width = 30, 
           height = 5, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name ="BuPu")))(100),
           scale='column',
           annotation_colors = annColors)

  pheatmap(plot_de_100, 
         annotation_row = my_pheno_col, 
         cluster_rows = FALSE, gaps_row = 26, 
         filename = 'degene_scaled_BuPu_100_LineNames_tall.pdf', 
         width = 30, 
         height = 12.5, 
         color = colorRampPalette(rev(brewer.pal(n = 7, name ="BuPu")))(100),
         scale='column',
         annotation_colors = annColors)

 
```


# HOUSEKEEPING
```{r}

sessionInfo()
setwd(pathToRData)
save.image(file = paste0(currDate,'_',expName,'_allVars.RData'))


```













